The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-05-29 source/sec_SGD.tex
minor changes

	With full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}=O(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}\leq O(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}
\end{theorem}
%
\begin{remark}
	\textbf{Linear speedup. }We first compare our bound with that in {[}ICLR{]},
	which is $O(\frac{1}{NT}\nu_{\max}^{2}\sigma^{2}\kappa/\mu+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{E^{2}G^{2}}{T}\kappa/\mu)$.
	Because the third term $\frac{E^{2}G^{2}}{T}\kappa/\mu$ is also $O(1/T)$
	without a dependence on $N$, for any choice of $E$ their bound cannot
	achieve linear speedup. The improvement of our bound comes from the
	term $\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}$, which now is $O(1/T^{2})$.
	As a result, all leading terms scale with $1/N$ in the full device
	particiaption setting, and with $1/K$ in the partial participation
	setting. This implies that in both settings, there is a linear speedup
	in the number of active workers during the communication round.
\end{remark}
%
\begin{remark}
	\textbf{Choice of $E$.} Our bound implies a distinction in the choice
	of $E$ between the full and partial participation settings. In full
	participation, as long as $E=O(\sqrt{\frac{T}{N}})$, the term $\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}=O(\frac{1}{NT})$,
	which is on the same order as the leading term $\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}$.
	Thus to achieve a linear speedup in the number of workers $N$, it
	suffices to communicate every $O(\sqrt{\frac{T}{N}})$ iterations.
	In contrast, the bound in {[}ICLR{]} does not allow $E$ to scale
	with $\sqrt{T}$, even for full participation. Thus our bound yields
	a more efficient communication complexity, while at the same time
	providing a linear speedup in the convergence. On the other hand,
	with partial participation, the term $\frac{\kappa E^{2}G^{2}/\mu}{KT}$
	is also a leading term, and so $E$ must be $O(1)$. In this setting,
	our bound still yields a linear speedup in $K$, which is not the
	case for previous analyses. 
\end{remark}
%
\textbf{}%
\begin{comment}
This implies that $E$ canot be chosen $O(T^{\beta})$ for any $\beta>0$
without degrading the performance. This should be checked in experiments,
whether with partial participation if the communication round is set
to scale with $T$, the convergence deteriorates. This is in constrast
with the full participation case, where $E=O(\sqrt{\frac{T}{N}})$
is allowed. \textbf{If we can confirm that full participation allows
linear speedup with $\nu=N\cdot\max_{k}p_{k}\approx1$ and $E=O(\sqrt{\frac{T}{N}})$,
whereas partial participation only allows $E=O(1)$, then this would
an interesting phenomenon that is not reported by previous studies!}

The convergence result implies that with fixed $E$, as long as \$N\$
satisfies \$E=O(\textbackslash sqrt\{T/N\})\$, there is linear speedup.
When \$N\$ exceeds an upper bound, however, linear speedup may fail
to happen, and \$T/N\$ may remain constant, i.e. the number of iterations
required for convergence may increase with \$N\$. 
\end{comment}

>>>>

	With full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}=O(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}\leq O(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}})
	\end{align*}
\end{theorem}
%
\begin{remark}
	\textbf{Linear speedup. }We first compare our bound with that in {[}ICLR{]},
	which is $O(\frac{1}{NT}\nu_{\max}^{2}\sigma^{2}\kappa/\mu+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{E^{2}G^{2}}{T}\kappa/\mu)$.
	Because the third term $\frac{E^{2}G^{2}}{T}\kappa/\mu$ is also $O(1/T)$
	without a dependence on $N$, for any choice of $E$ their bound cannot
	achieve linear speedup. The improvement of our bound comes from the
	term $\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}$, which now is $O(1/T^{2})$.
	As a result, all leading terms scale with $1/N$ in the full device
	particiaption setting, and with $1/K$ in the partial participation
	setting. This implies that in both settings, there is a linear speedup
	in the number of active workers during the communication round.
\end{remark}
%
\begin{remark}
	\textbf{Choice of $E$.} Our bound implies a distinction in the choice
	of $E$ between the full and partial participation settings. In full
	participation, as long as $E=O(\sqrt{\frac{T}{N}})$, the term $\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}=O(\frac{1}{NT})$,
	which is on the same order as the leading term $\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}$.
	Thus to achieve a linear speedup in the number of workers $N$, it
	suffices to communicate every $O(\sqrt{\frac{T}{N}})$ iterations.
	In contrast, the bound in {[}ICLR{]} does not allow $E$ to scale
	with $\sqrt{T}$, even for full participation. Thus our bound yields
	a more efficient communication complexity, while at the same time
	providing a linear speedup in the convergence. On the other hand,
	with partial participation, the term $\frac{\kappa E^{2}G^{2}/\mu}{KT}$
	is also a leading term, and so $E$ must be $O(1)$. In this setting,
	our bound still yields a linear speedup in $K$, which is not the
	case for previous analyses. 
\end{remark}
